---
title: "factorial2x2: Design and Analysis of a 2x2 Factorial Trial with a Time-to-Event Endpoint"
author: "Eric Leifer and James Troendle"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Motivation for the factorial2x2 package

The two-by-two factorial design  randomizes subjects to receive treatment either $A$ alone, treatment $B$ alone, both treatment $A$ and $B$ $(AB)$, or neither treatment $(C)$.     When the combined effect of $A$ and $B$ is less than the sum of the separate $A$ and $B$ effects, called *subadditivity*, there can be low power to detect the $A$ effect using an overall  test which compares the $A$ and $AB$ groups to the $C$ and $B$ groups.  In such an instance, simple effect tests such as $A$ vs. $C$ and $AB$ vs. $C$ may be useful since they are not affected by subadditivity.  However, the simple effect tests  can  have low power since they use half the subjects of the overall test.

Consider the $A$ research question which examines the benefit of $A$ either by itself, or in combination with $B$. To exploit the sample size advantage of the overall test and robustness to subadditivity of the simple tests,
various combinations of overall and simple effects tests may be considered.
We consider three multiple testing procedures:

1.  The **2/3-1/3 procedure** which allocates 2/3 of the significance level to testing the overall $A$ effect and 1/3 of the significance level to the simple $AB$ effect.

2. The **1/3-1/3-1/3 procedure** which allocates 1/3 of the significance level to testing each of the overall $A$, simple $A$, and simple $AB$ effects.

3.  The **1/2-1/2 procedure** which allocates 1/2 of the significance level to testing the simple $A$ effect and 1/2 to the simple $AB$ effect.


`factorial2x2` has two main R functions:  `fac2x2analyze` computes the hazard ratio, 95% confidence interval, and nominal p-value for the overall $A$, simple $A$, and simple $AB$ hazard effects.  It  also performs signficance testing
for the three multiple testing procedures.
`fac2x2design` calculates the power for the overall and simple tests as well as the three multiple testing procedures.



## Example using fac2x2analyze 

We use the simulated data in `simdata` which are data (included with the `factorial2x2` package) from a simulated 2x2 factorial clinical trial of 4600 subjects. Subjects are simultaneously randomized to receive either treatment $A$ or placebo as well as treatment $B$ or placebo.
We are interested in testing and estimating the overall $A$ effect, the simple $A$ effect, and the simple $AB$ effect.  We are also interested in the three multiple testing procedures described in the previous section.  Below is the R code and output for the three testing procedures.  Based on the below results, the 2/3-1/3 procedure detects the simple *AB* effect, the 1/3-1/3-1/3 procedure detects the simple *A* and *AB* effects, and the 1/2-1/2 procedure detects the simple *A* and simple *AB* effects.

```{r, eval = F, echo = F, results = "hide"}
# When I use knitr to run this whole file, I need to 
# "require" all the packages that are needed to run the code
# since knitr uses a new environment.  This is in contrast to
# running the code chunks separately which use all of the packages
# which are listed in the factorial2x2 package's DESCRIPTION file
# on the DEPENDS line

# require(factorial2x2)
# require(survival)
# require(mvtnorm)
```

```{r, eval =   F}
time <- simdata[, 'time']    # follow-up time
event <- simdata[, 'event']  # event indicator
indA <- simdata[, 'indA']    # treatment A indicator
indB <- simdata[, 'indB']    # treatment B indicator
test <- fac2x2analyze(time, event, indA, indB, simdata[, 6:10], niter = 5)
# simdata[, 6:10] corresponds to the adjustment variables which include
# a history of cardiovascular disease and (yes/no) and four indicator
# variables which correspond to which of 5 clinical centers enrolled the patient
test
$hrA  
[1] 0.8895135          # overall A effect HR

$ciA
[1] 0.786823 1.005607  # 95% CI for overall A effect HR

$pvalA
[1] 0.06139083         # p-value for overall A effect HR

$hra
[1] 0.8096082          # simple A effect HR

$cia
[1] 0.6832791 0.9592939 # 95% CI for simple A effect HR

$pvala
[1] 0.01468184         # p-value for simple A effect HR

$hrab
[1] 0.7583061          # simple AB effect HR

$ciab
[1] 0.6389355 0.8999785 # 95% CI fo simple A effect HR

$pvalab
[1] 0.001545967       # p-value for simple AB effect HR

$sig23A
[1] 0.03333333        # significance criterion for overall A effect for 2/3-1/3 procedure

$sig23ab
[1] 0.0256049         # signficance criterion of simple AB effect for 2/3-1/3 procedure

$result23
[1] "accept overall A" "reject simple AB"   # hypothesis testing results for 2/3-1/3 proceudre

$sig13
[1] 0.02091404        # significance criterion for all effects for 1/3-1/3-1/3 procedure

$result13
[1] "accept overall A" "reject simple A"  "reject simple AB"  # hypothesis testing results

$sig12
[1] 0.02665078       # significance criterion all effects for 1/2-1/2 procedure

$result12
[1] "reject simple A"  "reject simple AB"   # hypothesis testing results
```

## Example using fac2x2design

Here we use fac2x2design to compute the power for Scenario 5 in Table 2 from Leifer, Troendle, et al. (2019).

```{r, eval =   F}
  n <- 4600          # total sample size
  rateC <- 0.0445    # one year event rate in the control group
  hrA <- 0.80        # simple A effect hazard ratio
  hrB <- 0.80        # simple B effect hazard ratio
  hrAB <- 0.72       # simple AB effect hazard ratio
  mincens <- 4.0     # minimum censoring time in years
  maxcens <- 8.4     # maximum censoring time in years
  
  fac2x2design(n, rateC, hrA, hrB, hrAB, mincens, maxcens, dig = 2, alpha = 0.05)
  $powerA
  [1] 0.7182932      # power to detect the overall A effect at the two-sided 0.05 level
 
  $power23.13
  [1] 0.9290271      # power to detect the overall A or simple AB effects using the 
                     # 2/3-1/3 procedure
 
  $power13.13.13
  [1] 0.9302084      # power to detect the overall A, simple A, or simple AB effects using 
                     # the 1/3-1/3-1/3 procedure
 
  $power12.12
  [1] 0.9411688      # power to detect the simple A or simple AB effects using the 
                     # 1/2-1/2 procedure
```


## References
Leifer, E.S., Troendle, J.F., Kolecki, A., Follmann, D. Joint testing of overall and simple effect for the two-by-two factorial design. 2019. Submitted.

Slud, E.V. Analysis of factorial survival experiments. Biometrics. 1994; 50: 25-38.
